### 2015-09-12 manuscript.Rnw
first editorial pass at introduction

\begin{center}
  \noindent{\Large \bf \code{rotl} an R package to interact with the Open Tree
    of Life Data}

@@

\noindent{\normalsize \sc Fran\c{c}ois Michonneau}\\
\noindent {\small \it $^1$Florida Museum of Natural History & Whitney Laboratory
  for Marine Sciences, University of Florida, Gainesville,
  FL 32611-7800, USA}\\
\end{center}
\medskip

@@

Phylogenetic information has become central to many facets of the life
sciences. However, the information is scattered across many publications, often
only available as image files and the variety of file formats used to store
phylogenetic information makes it difficult for researchers outside the field of
phylogenetics to access this information.

The Open Tree of Life (http://opentreeoflife.org) aims at assembling and
synthetizing our current understanding of the phylogenetic relationships across
all organisms on Earth \cite{Hinchliff2015}. This aims is achieved by combining
taxonomic information with published phylogenies. This is made possible by
mapping species found in published studies to species in the taxonomy, a
taxonomic lookup service is therefore needed. Presentation of the OTL:
synthesis, 2.3M tips, continuously updated.

This resource has many applications across the life sciences to explore
fundamental questions in evolution, biodiversity, conservation, and has great
potential for education and outreach purposes.

OTL also provides web interface to explore the open tree (and download
subtrees?)

Open Tree of Life provides an Application Programming Interface (API) to the
different services it offers. Here, we present \code{rotl} an R pacakge that
allows users to interact with the OTL's API to import phylogenetic trees and
taxonomic information in R where this information can be used to be analyzed
with the rich suites of packages available for phylogenetic analyses.

R has become a popular software and programming environment to analyze, explore
and visualize phylogenetic data because of the variety of packages available
\cite{}. ape, phytools, geiger, ggtree, phylobase, etc... (cite task view).

The OTL's API, and in turn \code{rotl}, provides services (1) to download a tree
from a set of species names, this can be useful to illustrate a manuscript, map
the evolution of a character in a taxonomic group, etc....; (2) to download a
tree published in a paper that might be used to build the synthetic tree, this
can be useful to reproduce or expand on an already published analysis; (3) to
download information about the taxonomy across all organisms considered by OTL
in a phylogenetic context.

Many other applications, getting trees for educational purpose (illustrating
course material, explore relationships among species), illustrate manuscript.

\section{General features}

The tree can be imported into R memory to interact with other packages and be
used in comparative analyses, or they can be written to files so they can
interact with other programs.

@@

@@

No branch lengths for trees from the synthetic TOL, limit applications for
comparative data, but could be possible to use calibrations from time tree
databases.

@@

\code{rotl} is available both from the CRAN repository
(https://cran.r-project.org/package=rotl) and from GitHub
(https://github.com/ropensci/rotl).

>>>>

\begin{center}
  \noindent{\Large \bf \texttt{rotl} an R package to interact with the Open Tree
    of Life Data}

@@

\noindent{\normalsize \sc Fran\c{c}ois Michonneau}\\
\noindent {\small \it $^1$Florida Museum of Natural History \& Whitney Laboratory
  for Marine Sciences, University of Florida, Gainesville,
  FL 32611-7800, USA}\\
\end{center}
\medskip

@@

Phylogenetic information data is being used in many disciplines of the
life sciences, and phylogenetic comparative methods have seen their
use diversified in recent years (community phylogenetics,
...). However, phylogenetic information is scattered, often only
available as image files within publications \cite{Mounce}, and the
variety of file formats used to store phylogenetic information makes
it difficult for researchers to access

The Open Tree of Life (http://opentreeoflife.org) aims at assembling
and synthesizing the current understanding of phylogenetic
relationships across all organisms on Earth \cite{Hinchliff2015},
while providing tools and services to allow users to access this
information easily. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published
phylogenies to elucidate relationships among taxa.

The current draft of the Open Tree of Life contains 2.3 million
tips. Because of its comprehensive nature, researchers have access to
phylogenetic information for species that have yet to be included in
any phylogenetic tree as their position is inferred based on
taxonomy. Beyond obvious applications across the life sciences to
explore research questions in evolution, biodiversity, and
conservation, the resources OTL provides could be used in education
and outreach (e.g., illustrating course material, develop outreach
activities to explore relationships among species).

The Open Tree of Life project provides a web interface to explore,
query and download parts of the Open Tree
(http://tree.opentreeoflife.org). Additionally, the project provides
an Application Programming Interface (API) to the different services
it offers allowing users to interact programmatically with the
data. Here, we present \texttt{rotl} an R package that allows users to
interact with the OTL's API to search for and import phylogenetic
trees and taxonomic information. Having access to this data directly
from within R allows researchers to take advantage of the multiple R
packages that are available to visualize, analyze and manipulate
phylogenetic trees (ape, phytools, geiger, ggtree, phylobase,
etc... cite task view).

\section{Presentation of the services provided by OTL}

Presentation of the OTL: synthesis, 2.3M tips, continuously
updated. There are therefore 4 resources that the OTL provides: (1)
the taxonomic information used as the backbone from the tree, (2) the
trees and associated published studies that are used to build the
synthetic tree, (3) a taxonomic name resolution service used to map
species names to the Open Tree Taxonomy, (4) the synthetic tree. Also,
Graph of Life endpoint but not of use for normal users and beyond the
scope of this paper.

\texttt{rotl} provides access to these services so the data stored in
OTL (including trees) can be imported into R (either in memory or by
writing to files) to interact with other packages and be used in
comparative analyses, or they can be written to files so they can used
by other programs.

@@

The OTL's API, and in turn \texttt{rotl}, provides services (1) to
download a tree from a set of species names, this can be useful to
illustrate a manuscript, map the evolution of a character in a
taxonomic group, etc....; (2) to download a tree published in a paper
that might be used to build the synthetic tree, this can be useful to
reproduce or expand on an already published analysis; (3) to download
information about the taxonomy across all organisms considered by OTL
in a phylogenetic context.

Tree versionning? Replicability of analyses

@@

No branch lengths for trees from the synthetic TOL, limit applications for
comparative data, but could be possible to use calibrations from time tree
databases. Or combined with partial chronograms and use \cite{phyndr}.

@@

\texttt{rotl} is available both from the CRAN repository
(https://cran.r-project.org/package=rotl) and from GitHub
(https://github.com/ropensci/rotl).
